---
title: "PBD_ABC_progress"
author: "Thijs Janzen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RcppParallel)
library(pbdABC)
library(tidyverse)
```

# Progress PBD-ABC

Steps:

1)	Be able to simulate Ltables using PBD
2)	Calculate gamma and colless directly from Ltable
3)	Set up multi-threaded ABC
4)	Verify multi-threaded ABC
5)	Do PBD-ABC

## Step 1: create valid Ltables
PBD simulations are implemented in C++, previously verified for the treestats paper. They return Ltables including extinct species. Unfortunately, we are only interested in reconstructed trees. This is especially important for the Colless statistic (it is easy to remove those branching points leading to extinct species to calculate the gamma statistic).

Thus, I have designed an algorithm to remove extinct species from an Ltable. To do so, first, all extinct entries are removed from the Ltable. However, this may break the Ltable, as some parents are now removed. 
Thus, we loop over the Ltable, and for those species that have a parent that no longer exists in the Ltable, we look up the parent in the original Ltable (the one that included extinct species), and then replace the orphan species with the information from the extinct parent. We do this in reverse order, e.g. starting at the youngest species. Thus, the replaced orphan species is checked again, and in case that itself also has a parent that does not exists, it gets replaced again.

Now we have a functional Ltable where all species have existing parents. However, numbering is most likely not correct, and the code corrects for this to ensure that the first species has the absolute ID of 1, the second 2 etc.

To verify this, a test function returns both the raw Ltable, and the dropped Ltable. When converting these to a phylogeny, they should yield the same resulting phylogeny.
In these simulations, for sake of simplicity, I have disabled protracted speciation (e.g. setting the completion rate to infinity, and setting l1 and m1 to zero).

```{r test ltable}

for (r in 1:10) {
  sim_result <- pbdABC::test_simulations(birth = 1, death = 0.5, crown_age = 5)
  tree1 <- treestats::l_to_phylo(sim_result$ltable,
                                 drop_extinct = TRUE)
  tree2 <- treestats::l_to_phylo(sim_result$ltable_dropped,
                                 drop_extinct = TRUE)
  s1 <- treestats::calc_all_stats(tree1)
  s2 <- treestats::calc_all_stats(tree2)
  testthat::expect_equal(s1, s2)
}
```

No errors appear, indicating that the method is accurate.

## Step 2: Calculate Gamma and Colless accurately from an Ltable

The treestats package can be included as an Rcpp header, and then the functions inside it can be accessed directly. This allows for calculation of the Gamma and Colless statistics *without* going back to R. However, it seems prudent to verify that these calculations are correct and that we are passing everything correctly. Again, we can use the testing function, as this also returns the calculated Gamma and Colless values (on the reconstructed Ltable).

```{r test stats}
for (r in 1:10) {
  sim_result <- pbdABC::test_simulations(birth = 1, death = 0.5, crown_age = 5)
  tree1 <- treestats::l_to_phylo(sim_result$ltable, drop_extinct = TRUE)
  tree2 <- treestats::l_to_phylo(sim_result$ltable_dropped,
                                 drop_extinct = TRUE)
 
  g1 <- treestats::gamma_statistic(tree1)
  g2 <- treestats::gamma_statistic(tree2)
  g3 <- treestats::gamma_statistic(sim_result$ltable_dropped)
  g4 <- sim_result$gamma
  testthat::expect_equal(g1, g2)
  testthat::expect_equal(g1, g3)
  testthat::expect_equal(g1, g4)

  c1 <- treestats::colless(tree1)
  c2 <- treestats::colless(tree2)
  c3 <- treestats::colless(sim_result$ltable_dropped)
  c4 <- sim_result$colless
  testthat::expect_equal(c1, c2)
  testthat::expect_equal(c1, c3)
  testthat::expect_equal(c1, c4)
}
```

## Step 3/4: Set up multi-threaded ABC-SMC

ABC_SMC can be quite computationally intensive. However, because particles for
each iteration are independent, they can be generated in parallel.
I have extended the Rcpp code using the RcppParallel package to make use of
multithreading. Before letting the package loose on a PBD tree, I think it
makes sense to first verify everything on a B-D tree. 
Using a simple Yule tree, I set out to accurately reproduce the 'true' birth
rate. This turned out to be rather complicated, due to a number of errors from
my part: 1) rather than comparing to the true birth rate, the aim is to find
the ML birth rate (which may differ a bit from the true rate), 2) calculation
of the weights used for resampling in the SMC procedure was flawed, which caused
sometimes to generate NaN values, messing up estimations, 3) I incorrectly 
thought the prior was biasing estimates (where this in reality was due to issues
1 and 2), and implemented two different possible priors. Thus, we can now compare
the impact of these priors - which should be negligible ideally, given not too
small trees. 

ABC is performed using the Colless and Gamma statistic, with an exponentially
reducing threshold, and a perturbation kernel scaled to the variance of the 
posterior of the previous iteration. I used 1000 particles per generation. 
Estimations were stopped when the acceptance rate fell below 1e-6, or when 
5 hours of walltime (with 10 threads) were used up. If we want to push further,
we could extend this of course, but for this exercise, this should suffice.

### Priors
I made use of two priors:
1) an exponential prior with rates: c(1, 10, 10000, 10000, 0.01)
for lambda0, mu0, lambda1, mu1 and the completion rate. This does help the estimate a bit,
as we by default estimate lambda1 and mu1 to be very low.
2) a log-uniform prior, which is uniform in 10^(lower, upper). I picked for lower:
c(-5, -5, -5, -5, -5) and for upper: c(3, 3, 3, 3, 5). 

Please note that the prior is used to generate the initial generation of particles,
but is also later used to calculate weights, and to reject particles if they are
generated outside of the prior.

### Yule tree

I performed 30 replicate simulations of Yule trees with a birth rate of 1, aiming
for trees with 150 tips, using a crown age of: T = (1/b) * log(N/2), with b the
birth rate (of 1) and N the target number of tips. This yields a crown age of
4.3, which should give trees with an expected size of 150. For every tree, I also
inferred the birth rate using DDD::bd_ml, and compare the ABC estimate with that.


```{r load data, include=FALSE}
exp_found <- readRDS("/Users/thijsjanzen/pbd_abc_yule_exp_prior.rds")
log_found <- readRDS("/Users/thijsjanzen/pbd_abc_yule_loguniform_prior.rds")

exp_found$prior <- "exponential"
log_found$prior <- "log_uniform"

found <- rbind(exp_found, log_found)
```
```{r plot_yule}
found %>%
  gather(key = "to_plot", value = "val", c(d_gamma, d_colless, d_num_lin)) %>%
  ggplot(aes(x = iter, y = val, group = iter)) +
    geom_boxplot(outlier.shape = NA) +
    geom_hline(yintercept = 0.0) +
    theme_classic() +
    theme(legend.position = "none") +
    facet_grid(rows = vars(to_plot), cols = vars(prior), scales = "free")

found %>%
  gather(key = "to_plot", value = "val", c(d_lambda0)) %>%
  ggplot(aes(x = iter, y = val, group = iter)) +
    geom_boxplot(outlier.shape = NA) +
    geom_hline(yintercept = 0.0) +
    theme_classic() +
    theme(legend.position = "none") +
    facet_grid(rows = vars(to_plot), cols = vars(prior), scales = "free")
```

Here, we see two interesting patterns: Using either prior, the ABC algorithm
hones in rapidly to obtain gamma and colless values that are close to the observed
values (the top plot plots the difference with the true value, centered around 0).

Then, the bottom plot shows us the fit to the ML estimate of lambda. We see
that using the exponential prior, the estimate is spot on.
Using the log-uniform prior, the estimate is slightly below the ML value, and
for the last few iterations seems to deviate quite a bit, possibly due to a lower
number of replicates reaching this many iterations.

### Birth-Death tree

Here, we use a birth rate of 0.75, and a death rate of 0.25. The crown age is
now 8.63

```{r load data bd, include=FALSE}
exp_found <- readRDS("/Users/thijsjanzen/pbd_abc_bd_exp_prior.rds")
log_found <- readRDS("/Users/thijsjanzen/pbd_abc_bd_loguniform_prior.rds")

exp_found$prior <- "exponential"
log_found$prior <- "log_uniform"

found <- rbind(exp_found, log_found)
```
```{r plot_bd}
found %>%
  gather(key = "to_plot", value = "val", c(d_gamma, d_colless, d_num_lin)) %>%
  ggplot(aes(x = iter, y = val, group = iter)) +
    geom_boxplot(outlier.shape = NA) +
    geom_hline(yintercept = 0.0) +
    theme_classic() +
    theme(legend.position = "none") +
    facet_grid(rows = vars(to_plot), cols = vars(prior), scales = "free")

found %>%
  #gather(key = "to_plot", value = "val", c(d_lambda0 - d_mu0)) %>%
  ggplot(aes(x = iter, y = d_div, group = iter)) +
    geom_boxplot(outlier.shape = NA) +
    geom_hline(yintercept = 0.0) +
    theme_classic() +
    ylim(-2, 2) + 
    theme(legend.position = "none") +
    facet_wrap(~prior)
```

Here, it seems like the exponential prior yields a slight overestimate of the
diversification rate, whereas the log-uniform prior seems on par. However, this 
is a bit misleading, as many outliers have been removed here for the sake of
visualisation. Furthermore, it appears that the priors have influenced the result
a bit, as for the log-uniform prior, the extinction rate could not drop below
1e-5 - and in general, the model tried to force the extinction rate to zero, and
estimate birth rate of 0.5:

```{r plot_bd_dens}
found %>%
  filter(iter == 13) %>%
 ggplot(aes(x = mu0, color = prior)) +
    geom_density() +
    theme_classic() +
    scale_x_log10()

found %>%
  filter(iter == 13) %>%
 ggplot(aes(x = lambda0, color = prior)) +
    geom_density() +
    theme_classic() +
    scale_x_log10()
```

 

